## Example of a prediction model for significant wave height (SWH) based on sea level pressure (SLP)
### Authors: Hugo L. Hammer, Tor Arne Oigaard, Thordis Thorarinsdottir and Hanne Rognebakke

Scripts are located in the folder script.
Data example is located in the folder data. 

R working directory should be set to the location of the script folder.

To run the analysis, run the script run_models.r as shown below. 

Load necessary packages and source functions.
```{r message=FALSE}
require(glmnet)
require(forecast)
source("getPreddistr.r")
source("BoxCoxLambda.r")
source("BoxCoxLambdaKnown.r")
source("fourier.r")
source("qBoxCox.r")
source("pBoxCox.r")
```

Load ERA-Interim data 
```{r}
load("../data/SWH_SLP_SLP.grad.subset.7.Rdata")
longitudeSWH
latitudeSWH
range(years.all)
summary(SWH)
summary(SLP)
summary(SLP.grad)
```

Set which grid cell to model (out of a total of 7x7 cells).
```{r}
lonSWH = 4
latSWH = 4
```

Define how many missing values is "good enough", and define the model to use.
```{r}
na.thresh = 500
spatial.neighborhoods = 2 
models.fourier = c("Proposed model, AR order = 5, spatial neigb. = ", spatial.neighborhoods, "\n")
n.models.fourier = length(models.fourier)
```

Create Fourier terms to use in the modeling
```{r}
m.fourier = 365.25*4 # Number of observations per year. Four observations per day.
K.fourier = 2 # Number of Fourier terms
n.time = length(time.all)
intercept.fourier = fourier(rep(1,n.time), K=K.fourier, m=m.fourier)
colnames(intercept.fourier) = paste("intercept", colnames(intercept.fourier), sep = "_")
```

Split that data in a training set and a test set 
```{r}
training.test = list()
training.test[[1]] = which(years.all %in% 2006:2014)
training.test[[2]] = which(years.all %in% 2015)
```

Estimate model parameters and obtain predictive distributions
```{r}
pred.dist  <- getPreddistr(j=lonSWH, k=latSWH, neig=spatial.neighborhoods)
pred.mean = pred.dist$pred.mean
pred.sd = pred.dist$pred.sd
pred.lambda = pred.dist$pred.lambda
print(pred.dist$fits)
```

Create the vector of observations in the test period
```{r}
obs  <- SWH[lonSWH, latSWH, training.test[[2]]]
```

Calculate PIT values and plot a PIT histogram for the test set 
```{r}
pit  <- pBoxCox(obs, pred.mean, pred.sd, pred.lambda)
hist(pit, freq=FALSE, nclass=10, col="gray", xlab="PIT value", 
     main=paste("Lon = ", longitudeSWH[lonSWH], ", Lat = ", latitudeSWH[latSWH]))
abline(a=1, b=0, lty=2)
```

Calculate the mean absolute error over the test period
```{r}
pred.median  <- qBoxCox(0.5, pred.mean, pred.sd, pred.lambda)
mae  <- mean(abs(obs - pred.median))
mae
```
Plot the prediction and the observation in the first and last 100 time points of the test period
```{r}
upper  <- qBoxCox(0.95, pred.mean, pred.sd, pred.lambda)
lower  <- qBoxCox(0.05, pred.mean, pred.sd, pred.lambda)

t.period  <- c(1:100)
t.upper  <- upper[t.period]
t.lower  <- lower[t.period]

plot(t.period, obs[t.period], type="l",
     xlab="Time point in test period", ylab="SWH", ylim=c(0,15),
     main=paste("Lon = ", longitudeSWH[lonSWH], ", Lat = ", latitudeSWH[latSWH]))
polygon(c(t.period, rev(t.period), t.period[1]),
        c(t.lower, rev(t.upper), t.lower[1]), col="gray90", border="NA")
lines(t.period, obs[t.period])
lines(t.period, pred.median[t.period], col="gray50", lty=2)
legend("topright", lty=c(1,2,1), lwd=c(1,1,4), col=c("black", "gray50", "gray90"),
       legend=c("Observation", "Predictive median", "90% prediction interval"))

t.period  <- c(1361:1460)
t.upper  <- upper[t.period]
t.lower  <- lower[t.period]

plot(t.period, obs[t.period], type="l",
     xlab="Time point in test period", ylab="SWH", ylim=c(0,15),
     main=paste("Lon = ", longitudeSWH[lonSWH], ", Lat = ", latitudeSWH[latSWH]))
polygon(c(t.period, rev(t.period), t.period[1]),
        c(t.lower, rev(t.upper), t.lower[1]), col="gray90", border="NA")
lines(t.period, obs[t.period])
lines(t.period, pred.median[t.period], col="gray50", lty=2)
legend("topright", lty=c(1,2,1), lwd=c(1,1,4), col=c("black", "gray50", "gray90"),
       legend=c("Observation", "Predictive median", "90% prediction interval"))
```
